Assessment and implementation of resource road simulation methods for projection of cumulative effects of disturbance on wildlife

### 2023-01-17 analysis/paper/paper.Rmd

---
title: "Assessment and implementation of resource road simulation methods for projection of cumulative effects on wildlife"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: David Lapins
    email: david.lapins@pc.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Kyle Lochhead
    email: Kyle.Lochhead@gov.bc.ca
    institute: [FLNR]
    correspondence: false
  - name: Gregory Paradis
    email: gregory.paradis@ubc.ca
    institute: [UBC]
    correspondence: false

@@

The impacts of roads on wildlife are complex and varied. Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021;@trombulak_review_2000;@fahrig_effects_2009].    
In Canadian forests, there is ample evidence of linear features attracting predators such as black bears and wolves seeking food or easy movement, while repelling prey such as caribou and moose [@dickie_corridors_2020;@polfus_identifying_2011;@mumma_predation_2018;@beauchesne_disentangling_2013;@leblond_avoidance_2013;@laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011;@whittington_towns_2022]. Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter movement patterns [@tucker_moving_2018], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021;@apps_spatial_2013;@pinard_calving_2012;@fahrig_effects_2009]. Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020].

@@

Leston et al. (2020) project forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then add fire. This example highlights that roads are also important for harvest planning, and that road network projection methods are integrated into a widely used optimization-based spatial harvest planning tool. However, Patchworks is proprietary closed-source commercial software, and the road network projection algorithm contained therein is difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017]. Published methods intended to support strategic planning [@anderson_projecting_2004;@clark_three-stage_2000;@dean_finding_1997] require inputs that are not publicly available, and employ algorithms that are too complex and detailed for application at regional, provincial or national scales. More promising options for projecting forest road network development at large scales include an iterative least cost path method developed by Hardy [-@hardy_forest_2019], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022]. The former method is available as QGIS and Landis plugins, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate]. None of these methods have been formally assessed, compared, published, or implemented in a way that allows for easy integration into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016; @scheller_forest_2004].

In this work, we address the need for adequately accurate forest road network projection tools that are also widely and easily applicable by implementing simple iterative least cost path and minimum spanning tree methods in an open source R package. We compare the ability of these methods to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia using performance metrics relevant to wildlife. The adequacy of a road network projection method depends on the purpose of the projection, and it is important to select performance metrics that reflect the purpose. Common predictors in models of the impacts of roads on wildlife include distance to roads [], buffered disturbance footprint [@johnson_science_2020;@ibisch_global_2016], and density []. We expect projecting these measures to be less challenging than projecting the spatial location of each road segment, and we do not expect good projections of road locations from the algorithms considered here. Our goal is to assess the ability of simple iterative least cost path and minimum spanning tree methods to project footprint, distance and density metrics that are commonly used to predict impacts of roads on wildlife.

@@

<!-- TO DO: dynamic links to figs and tables. -->
We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because forestry is the dominant resource extraction activity in this region and much of the road network was built to access to forest harvest cutblocks (Figure 1). This allows projection of the forest road network from historic forest harvest data. Approximately 55% of the area is non-forested land where road building is difficult or impossible (e.g. alpine, lakes, swamp, brush, rock) [@brown_revelstoke_2010], so the area provides opportunities to observed the responses of road projection algorithms to barriers. Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministry_of_forests_revelstoke_2021]. The area contains wildlife species that are sensitive to roads (refs) including endangered southern mountain woodland caribou (refs).

@@

To spatially define the cost of building and traveling on a forestry road we developed a simple model informed by slope and major barriers (Table 1). Following the Interior Appraisal Manual of BC, we assume the base cost per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table 1). Existing roads are assigned a cost of 0. The cost raster consists of 1703936 pixels with a resolution of 100 meters by 100 meters.

@@

All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke TSA [@teucher_bcdata_2021].

@@

Simple methods for projecting low-cost routes from an existing road network to target locations (i.e. harvest cutblocks) include simple least cost paths (LCP), dynamic least cost paths (DLCP) and least cost path minimum spanning trees (MST) \@ref(fig:fig2)). We use Dijkstra's shortest path algorithm [-@dijkstra_note_1959] to determine the least cost path between a target location (e.g. centroid of a cutblock) and the existing road network. Least cost paths to each cutblock is determined independently in simple method (LCP). The dynamic least cost paths (DLCP) method evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths. Sequence is determined by a heuristic such as the distance to the nearest existing road in these examples. The minimum spanning tree (MST) method calculates the minimum cost network to connecting all cutblocks to the road network using Kruskal's algorithm [-@kruskal_shortest_1956]. See Appendix A for a more detailed explanation of these methods. We focus our analysis on the DLCP and MST approaches which yield more realistic road networks (Figure \@ref(fig:fig2)). We compared these methods to an alternative iterative least cost path method developed by Hardy [ref].

@@

The DLCP and MST methods connect target or landing locations within each cutblock to the existing road network, so an important first step is to select one or more landing locations in each block. We investigated the sensitivity of results to the pattern and density of landing locations by comparing results derived from the centroid of each cutblock to regular or randomly selected landing densities of 1 (low) and 10 (high) points per km^2^ (Figure \@ref(fig:fig2)). We used default settings in the Hardy QGIS plugin except for kidding distance which was set to 200m. A skidding distance of 200 m corresponds to \~ 9 pts/km^2^. We compared two resolutions of the cost surface by aggregating the 100m resolution cost raster to a coarser 1000 m resolution. In total, we compared 12 road projection method variants to a cutblocks only scenario with no projected roads (Table 2).

Table 2: Road projection method variants. 
```{r}
tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
MST|Regular|1|100 m and 1000 m
MST|Regular|10|100 m and 1000 m
MST|Random|1|100 m and 1000 m
MST|Random|10|100 m and 1000 m
MST|Centroid|N/A|100 m and 1000 m
DLCP|Regular|10|100 m 
Hardy QGIS|N/A|~9|100 m
Cutblocks only|N/A|N/A|100 m

@@

To make help make simple road network projection methods more easily and widely applicable, we implemented the simple least cost path, dynamic least cost path and minimum spanning tree methods in the open source R package 'roads'.

@@

```{r fig2, fig.cap="We compared minimum spanning tree (MST), dynamic least cost paths (DLCP) and least cost path (LCP) road network projection algorithms. Landing locations were selected as the centroid of the cutblock, or a random or regularly spaced sample of points inside each cutblock (Table 2). Target cutblocks are outlined in grey, existing roads are black, projected roads are yellow, observed roads are red. Cost (100 m resolution) is indicated by grey shading."}
knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
```

## Performance Metrics
In order to assess the performance of the projections, five characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, measured as whether or not roads are present in each pixel; 2) Road density, measured as density of roads in meters per hectare (m/ha); 3) Distance to nearest road, measured as distance from each raster cell to the nearest road; 4) Road disturbance footprint, measured with a 1 km buffer around roads [@ibisch_global_2016]; and 5) Forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks. The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020], and road density is is often used in studies of anthropogenic impacts on wildlife [@beazley_road_2004; more]. Forestry disturbance footprint and road density were calculated using the XX and rasterizeLineDensity functions in the `caribouMetrics` R package [@hughes]. Distance to nearest road was calculated using a fast parallel moving window method implemented in the getDistFromSource function of the `roads` package. Only the road and forestry disturbance footprints were calculated for the cutblocks only method since the other characteristics are not applicable without roads. All road characteristics were stored in raster format.

<!-- TO DO: update the details on getDistFromSource function - where is it now? -->
<!-- TO DO: Does the cutblocks only scenario include the old roads? If no, it should. And if so, it seems like the other metrics could be calculated in the cutblocks only scenario. It would be interesting to have the "no projection" baseline for all the metrics -->

The aspatial performance of the projections was assessed by calculating the mean values of each road network characteristic and comparing the projected and observed values. This was done across the TSA and within cutblocks by masking the road network characteristic rasters with cutblock polygons and calculating means within cutblocks. Spatial performance was assessed in two different ways. First, difference maps were created by subtracting the projected distance to the nearest road from the observed in order to highlight discrepancies between the road networks. The observed and projected road networks were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS. Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: 1) pre-existing footprint; 2) agree roadless, meaning that both projected and observed show no footprint; 3) agree roaded, meaning that both projected and observed show new footprint; 4) false positive, meaning the projection shows a footprint but no footprint was observed; and 5) false negative, meaning the projection shows no footprint but a footpint was observed. These categories were summarized for each projection and characteristic and used to calculate sensitivity, precision and F-measure of each metric [@lee_prediction_2021]. Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true. Sensitivity will be lower when observed roads are missed and precision will be lower when roads are projected that were not observed. F-measure attempts to balance sensitivity and precision to give an overall score.

@@

For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks or overall scales (Figure \@ref(fig:fig3)). All road network projection method variants accurately predict the forestry disturbance footprint, and slightly overproject the road disturbance footprint within cutblocks. The high density methods overproject the overall road disturbance footprint more than low density methods, and underproject overall distance to roads, but all method variants give reasonably good projections of overall distance to roads. All methods underproject road density, in part because within 100 m pixels all the methods generate straight road segments. All the methods also underproject road presence, and the problem is more severe for the low density method variants. In general, performance differences among low and high density method variants are larger than differences among projection algorithms. With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (Figure A-\@ref(fig:figA1)).

@@

@@

```{r fig3, fig.height=3, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 100 m resolution cost raster across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”). Values greater than 0 indicate overprojection of the metric and negative values indicate an underprojection. The cutblocks only method is not defined for the road density, road presence or distance to road characteristics."}
# organize data for ggplot
meanTable_long <- meanTable %>% dplyr::select(-runTime) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA NA klementQGIS",
                             "NA NA NA"),
                  labels = c("MST centroid", 
                             "Observed",
                             "MST random low density",
                             "MST regular low density",
                             "MST random high density",
                             "MST regular high density",
                             "DLCP regular high density",
                             "Hardy QGIS",
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

```{r fig5, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant using the 100 m resolution cost raster. F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted. Values closer to one indicate better performance."}

@@

<!-- Discussion pts
- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models)
- as expected, not great at projecting exact road locations. sources of discrepancy include switchback, stream crossings, inaccurate harvest locations. All of these could be improved with more detailed high resolution cost & harvest data.
- density of landing pts has a larger effect than choice of road projection algorithm. So maybe don't get too worked up about the differences between one algorithm and another. if more accurate projections are required, focus first on refining cost & harvest landing locations.
- Selected revelstoke because it is topographically complex, so a good challenge. However, This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong. The results would likely be less accurate if the landings were spread further apart. But also less topographically complex landscapes may be simpler to project.
- For some purposes, projections from relatively crude cost models may be adequate. For this example landscape projections of disturbance footprints and distance to road are pretty good (and better than nothing). Density is harder to project partly because none of these algorithms can project variation in density within 100 m pixels. If projection of impacts on wildlife is a goal, it may be helpful to consider including more easily projectable road predictors in wildlife response models (Bodner,). If interested in predictors that we did not consider, see example script. Can evaluate alternative metrics without rerunning the projections.
-->

@@

A mathematical weighted graph consisting of nodes and edges was used to parameterize either Kruskal's or Dijkstra's algorithms. When solving least cost paths using Dijkstra's algorithm, each node in the graph corresponded to a pixel in a raster of our study area (100m x 100m resolution). Whereas, in the graph used for solving the minimum spanning tree via Kruskal's algorithm, each node represented either a landing or a point on the existing road network. Therefore, a node can be represented by any one of three attributes in the forest road network: (i) a landing, (ii) a point on an existing road, or (iii) an intermediate point. The landing nodes represent log decks or sources of timber and the point on the existing road is the sink where the timber will end up. The intermediate points are plausible locations to construct a road and are thus free of barriers like major water bodies. The edges connecting the nodes follow an "octagonal" neighbourhood or allow a maximum of eight possible edges for a given node, with diagonal edges having a higher cost. Thus, our methods utilize a grid pattern which does not support curvature or switchbacks in road design as required in operational road design. Therefore, our road projection methods are for strategic modelling purposes only. Edges are assigned weights which penalize the algorithm to account for various impedance and road construction costs. When solving least cost paths, a cost surface raster was used. The pixels of the cost surface raster corresponded to nodes in the weighted graph and thus the edge weight was the average cost between the two nodes or pixels. When solving the minimum spanning tree, the edge weight corresponded to the total cost of the least cost path between the two nodes.

@@

```{r get-data2}
data_path_drvd <- "analysis/data/derived_data/TSA27_1000"

@@

```{r figA2, fig.height=4.5, fig.width=fig_widths["mid"], fig.cap="Comparison of performance metrics for each projection method variant for the fine resolutiom (100 m) and coarse resolution (1000 m) cost rasters. F measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}

>>>>

---
title: "Assessment and implementation of resource road simulation methods for projection of cumulative effects of disturbance on wildlife"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: David Lapins
    email: david.lapins@pc.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Kyle Lochhead
    email: Kyle.Lochhead@gov.bc.ca
    institute: [FLNR]
    correspondence: false
  - name: Gregory Paradis
    email: gregory.paradis@ubc.ca
    institute: [UBC]
    correspondence: false

@@

The impacts of roads on wildlife are complex and varied. Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021;@trombulak_review_2000;@fahrig_effects_2009].    
In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus, Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020;@polfus_identifying_2011;@mumma_predation_2018;@beauchesne_disentangling_2013;@leblond_avoidance_2013;@laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011;@whittington_towns_2022; @proctor_effects_2020]. Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021;@apps_spatial_2013;@pinard_calving_2012;@fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020]. Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020].

@@

Leston et al. (2020) project forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then add fire. This example highlights that roads are also important for harvest planning, and that road network projection methods are integrated into a widely used optimization-based spatial harvest planning tool. However, Patchworks is proprietary closed-source commercial software, and the road network projection algorithm contained therein is difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017]. Published methods intended to support strategic planning [@anderson_projecting_2004;@clark_three-stage_2000;@dean_finding_1997] require inputs that are not publicly available, and employ algorithms that are too complex and detailed for application at regional, provincial or national scales. More promising options for projecting forest road network development at large scales include an iterative least cost path method developed by Hardy [-@hardy_forest_2019], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022]. The former method is available as QGIS and Landis plugins, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate]. These methods have not been formally assessed, compared, published, or implemented in a way that allows for easy integration into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016; @scheller_forest_2004].

In this work, we address the need for adequately accurate forest road network projection tools that are also widely and easily applicable by implementing simple iterative least cost path and minimum spanning tree methods in an open source R package. We compare the ability of these methods to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia using performance metrics relevant to wildlife. The adequacy of a road network projection method depends on the purpose of the projection, and it is important to select performance metrics that reflect the purpose. Common predictors in models of the impacts of roads on wildlife include distance to roads [], buffered disturbance footprint [@johnson_science_2020;@ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011], and density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017]. We expect projecting these measures to be less challenging than projecting the spatial location of each road segment, and we do not expect good projections of road locations from the algorithms considered here. Our goal is to assess the ability of simple iterative least cost path and minimum spanning tree methods to project footprint, distance and density metrics that are commonly used to predict impacts of roads on wildlife.

<!-- TO DO:
- more thorough review of SMC & grizzly papers
- summarize road metric types used in cited papers more systematically.
-->

@@

<!-- TO DO: dynamic links to figs and tables. -->
We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because forestry is the dominant resource extraction activity in this region and much of the road network was built to access to forest harvest cutblocks (Figure 1). This allows projection of the forest road network from historic forest harvest data. Approximately 55% of the area is non-forested land where road building is difficult or impossible (e.g. alpine, lakes, swamp, brush, rock) [@hachey_revelstoke_2010], so the area provides opportunities to observed the responses of road projection algorithms to barriers. Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministry_of_forests_revelstoke_2021]. The area is home to wildlife species that are sensitive to roads including grizzly bears and Southern Mountain woodland caribou [@hachey_revelstoke_2010]. Eight herds of Southern Mountain caribou have been extirpated or likely extirpated, and remaining herds are considered Endangered by COSEWIC and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020]. Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022]. Roads also alter grizzly habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].

@@

To spatially define the cost of building and traveling on a forestry road we developed a simple model informed by slope and major barriers (Table 1). Following the Interior Appraisal Manual of BC, we assume the base cost per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table 1). Existing roads are assigned a cost of 0. The cost raster consists of 1703936 pixels with a resolution of 1 ha.

@@

All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke Timber Supply Area [@teucher_bcdata_2021].

@@

Simple methods for projecting low-cost routes from an existing road network to target locations (i.e. harvest cutblocks) include simple least cost paths (LCP), dynamic least cost paths (DLCP) and least cost path minimum spanning trees (MST) \@ref(fig:fig2)). We use Dijkstra's shortest path algorithm [-@dijkstra_note_1959] to determine the least cost path between a target location (e.g. centroid of a cutblock) and the existing road network. Least cost paths to each cutblock is determined independently in simple method (LCP). The dynamic least cost paths (DLCP) method evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths. Sequence is determined by a heuristic such as the distance to the nearest existing road in these examples. The minimum spanning tree (MST) method calculates the minimum cost network to connecting all cutblocks to the road network using Kruskal's algorithm [-@kruskal_shortest_1956]. See Appendix A for a more detailed explanation of these methods. We focus our analysis on the DLCP and MST approaches which yield more realistic road networks (Figure \@ref(fig:fig2)). We compared these methods to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019].

@@

The DLCP and MST methods connect target or landing locations within each cutblock to the existing road network, so an important first step is to select one or more landing locations in each block. We investigated the sensitivity of results to the pattern and density of landing locations by comparing results derived from the centroid of each cutblock to regular or randomly selected landing densities of 1 (low) and 10 (high) points per km^2^ (Figure \@ref(fig:fig2)). We used default settings in the Hardy QGIS plugin except for kidding distance which was set to 200m. A skidding distance of 200 m corresponds to \~ 9 pts/km^2^. We compared two resolutions of the cost surface by aggregating the 1 ha resolution cost raster to a coarser 100 ha (1000 by 1000 m) resolution. In total, we compared 12 road projection method variants to a cutblocks only scenario with no projected roads (Table 2).

Table 2: Road projection method variants. 
```{r}
tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
MST|Regular|1|1 ha and 100 ha
MST|Regular|10|1 ha and 100 ha
MST|Random|1|1 ha and 100 ha
MST|Random|10|1 ha and 100 ha
MST|Centroid|N/A|1 ha and 100 ha
DLCP|Regular|10|1 ha 
Hardy QGIS|N/A|~9|1 ha
Cutblocks only|N/A|N/A|1 ha

@@

To make help make these road network projection methods more easily and widely applicable, we implemented the simple least cost path, dynamic least cost path and minimum spanning tree methods in the open source R package 'roads'.

@@

```{r fig2, fig.cap="We compared minimum spanning tree (MST), dynamic least cost paths (DLCP) and least cost path (LCP) road network projection algorithms. Landing locations were selected as the centroid of the cutblock, or a random or regularly spaced sample of points inside each cutblock (Table 2). Target cutblocks are outlined in grey, existing roads are black, projected roads are yellow, observed roads are red. Cost (1 ha resolution) is indicated by grey shading."}
knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
```

## Performance Metrics
In order to assess the performance of the projections, five characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, measured as whether or not roads are present in each pixel; 2) Road density, measured as density of roads in meters per hectare (m/ha); 3) Distance to nearest road, measured as distance from each raster cell to the nearest road; 4) Road disturbance footprint, measured with a 1 km buffer around roads [@ibisch_global_2016]; and 5) Forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks. The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020], and road density is is often used in studies of anthropogenic impacts on wildlife [@beazley_road_2004; @apps_factors_2006; more]. Forestry disturbance footprint and road density were calculated using the *disturbanceMetrics* and *rasterizeLineDensity* functions in the `caribouMetrics` R package [@hughes_cariboumetrics_2022]. Distance to nearest road was calculated using a fast parallel moving window method implemented in the *getDistFromSource* function of the `roads` package. Only the road and forestry disturbance footprints were calculated for the cutblocks only method since the other characteristics are not applicable without roads. All road characteristics were stored in raster format.

<!-- TO DO: Discuss why we do not calculate the 50m buffer metric used by Kyle for SMC. 50 m buffer is basically built in to the 100 m raster resolution. Road presence would be our closest equivalent measure - and we don't project that measure very well. -->

<!-- TO DO: update the details on getDistFromSource function - where is it, or where should it be? -->
<!-- TO DO: Does the cutblocks only scenario include the old roads? If no, it should. And if so, it seems like the other metrics could be calculated in the cutblocks only scenario. It would be interesting to have the "no projection" baseline for all the metrics -->

The aspatial performance of the projections was assessed by calculating the mean values of each road network characteristic and comparing the projected and observed values. This was done across the timber supply area and within cutblocks by masking the road network characteristic rasters with cutblock polygons and calculating means within cutblocks. Spatial performance was assessed in two different ways. First, difference maps were created by subtracting the projected distance to the nearest road from the observed in order to highlight discrepancies between the road networks. The observed and projected road networks were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS. Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: 1) pre-existing footprint; 2) agree roadless, meaning that both projected and observed show no footprint; 3) agree roaded, meaning that both projected and observed show new footprint; 4) false positive, meaning the projection shows a footprint but no footprint was observed; and 5) false negative, meaning the projection shows no footprint but a footpint was observed. These categories were summarized for each projection and characteristic and used to calculate sensitivity, precision and F-measure of each metric [@lee_prediction_2021]. Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true. Sensitivity will be lower when observed roads are missed and precision will be lower when roads are projected that were not observed. F-measure attempts to balance sensitivity and precision to give an overall score.

@@

For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks or overall scales (Figure \@ref(fig:fig3)). All road network projection method variants accurately predict the forestry disturbance footprint, and slightly overproject the road disturbance footprint within cutblocks. The high density methods overproject the overall road disturbance footprint more than low density methods, and underproject overall distance to roads, but all method variants give reasonably good projections of overall distance to roads. All methods underproject road density, in part because within 1 ha pixels all the methods generate straight road segments. All the methods also underproject road presence, and the problem is more severe for the low density method variants. In general, performance differences among low and high density method variants are larger than differences among projection algorithms. With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (Figure A-\@ref(fig:figA1)).

@@

<!-- TO DO: consider breaking down aspatial results into 'cutblocks', 'outside cutblocks' and 'overall': it may make the results easier to explain and interpret. -->

@@

```{r fig3, fig.height=3, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 1 ha resolution cost raster across the entire Revelstoke Timber Supply Area (“overall”) and within cutblocks (“cutblocks”). Values greater than 0 indicate overprojection of the metric and negative values indicate an underprojection. The cutblocks only method is not defined for the road density, road presence or distance to road characteristics."}
# organize data for ggplot
meanTable_long <- meanTable %>% dplyr::select(-runTime) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA NA klementQGIS",
                             "NA NA NA"),
                  labels = c("MST centroid", 
                             "Observed",
                             "MST random low density",
                             "MST regular low density",
                             "MST random high density",
                             "MST regular high density",
                             "DLCP regular high density",
                             "Hardy QGIS",
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

```{r fig5, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant using the 1 ha resolution cost raster. F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted. Values closer to one indicate better performance."}

@@

<!-- Discussion pts
- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models)
- as expected, not great at projecting exact road locations. sources of discrepancy include switchback, stream crossings, inaccurate harvest locations. Missing switchbacks and too many stream crossings also reduce our total amount of roads. All of these could be improved with more detailed high resolution cost & harvest and landing data. 
- density of landing pts has a larger effect than choice of road projection algorithm. So maybe don't get too worked up about the differences between one algorithm and another. if more accurate projections are required, focus first on refining cost & harvest landing locations.
- Selected revelstoke because it is topographically complex, so a good challenge. However, This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong. The results may be less accurate if the landings were spread further apart. But also less topographically complex landscapes may be simpler to project.
- For some purposes, projections from relatively crude cost models may be adequate. For this example landscape projections of disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing). Density is harder to project partly because none of these algorithms can project variation in density within 100 m pixels. If projection of impacts on wildlife is a goal, it may be helpful to consider including more easily projectable road predictors in wildlife response models (Bodner,). If interested in predictors that we did not consider, see example script. Can evaluate alternative metrics without rerunning the projections.
-->

@@

A mathematical weighted graph consisting of nodes and edges was used to parameterize either Kruskal's or Dijkstra's algorithms. When solving least cost paths using Dijkstra's algorithm, each node in the graph corresponded to a pixel in a raster of our study area (1 ha resolution). Whereas, in the graph used for solving the minimum spanning tree via Kruskal's algorithm, each node represented either a landing or a point on the existing road network. Therefore, a node can be represented by any one of three attributes in the forest road network: (i) a landing, (ii) a point on an existing road, or (iii) an intermediate point. The landing nodes represent log decks or sources of timber and the point on the existing road is the sink where the timber will end up. The intermediate points are plausible locations to construct a road and are thus free of barriers like major water bodies. The edges connecting the nodes follow an "octagonal" neighbourhood or allow a maximum of eight possible edges for a given node, with diagonal edges having a higher cost. Thus, our methods utilize a grid pattern which does not support curvature or switchbacks in road design as required in operational road design. Therefore, our road projection methods are for strategic modelling purposes only. Edges are assigned weights which penalize the algorithm to account for various impedance and road construction costs. When solving least cost paths, a cost surface raster was used. The pixels of the cost surface raster corresponded to nodes in the weighted graph and thus the edge weight was the average cost between the two nodes or pixels. When solving the minimum spanning tree, the edge weight corresponded to the total cost of the least cost path between the two nodes.

@@

<!-- TO DO: switch fig resolution labels from metres to ha. -->
```{r get-data2}
data_path_drvd <- "analysis/data/derived_data/TSA27_1000"

@@

```{r figA2, fig.height=4.5, fig.width=fig_widths["mid"], fig.cap="Comparison of performance metrics for each projection method variant for the fine resolutiom (1 ha) and coarse resolution (100 ha) cost rasters. F measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}
